---
title: "Reproducible Report on COVID19 Data"
author: "Thomas N"
date: "2023-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Loading Libraries Used
```{r load_libraries, message = FALSE}

library(tidyverse)
library(lubridate)
```

# Retrieving Data From Github and Reading It In
COVID-19 data was obtained from the the data repository by the Center for Systems Science and Engineering at Johns Hopkins University.  4 different files are used, which are: US confirmed cases, US deaths, global confirmed cases, and global deaths

```{r get_hopkins_data, message=FALSE}
url_in <- "https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series"

file_names <- 
c("time_series_covid19_confirmed_US.csv", 
  
"time_series_covid19_confirmed_global.csv", 
                
"time_series_covid19_deaths_US.csv", 

"time_series_covid19_deaths_global.csv")

urls <- str_c(url_in, file_names)
```


``` {r read_in_data, message = FALSE}
US_cases <- read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")

global_cases <- read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")

US_deaths <- read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")

global_deaths <- read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")


```

### Tidying Data
Remove columns unnecessary for this particular data analysis such as latitude and longitude.  Also converting column values to classes that are easier to work with to analyze, such as converting strings to date type.  Consolidating data into 2 separate data sets.  One that combines both cases and deaths within the US titled "US", and the the other that combines cases and deaths globally titled "global".

``` {r tidy_global_cases, message = FALSE}

global_cases <- global_cases %>%
  pivot_longer(cols = 
    -c('Province/State',
      'Country/Region', Lat, Long), names_to = "date", values_to = "cases") %>%
  select(-c(Lat,Long))

```


``` {r tidy_global_deaths, message = FALSE}

global_deaths <- global_deaths %>%
  pivot_longer(cols = 
    -c('Province/State',
      'Country/Region', Lat, Long), names_to = "date", values_to = "deaths") %>%
  select(-c(Lat,Long))
```


``` {r create_global, message = FALSE}

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
  Province_State = 'Province/State') %>%
  mutate(date = mdy(date))
```


``` {r tidy_global, message = FALSE}
global <- global %>% filter(cases >0)
#filters cases less than 0

#to check upper end of the cases
global %>% filter(cases > 100000000)

```

```{r tidy_US_cases, message = FALSE}
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
  names_to = "date",
  values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

```


```{r tidy_US_deaths, message = FALSE}
US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
  names_to = "date",
  values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

```

```{r tidy_US, message = FALSE}
US <- US_cases %>%
  full_join(US_deaths)

```


### Adding to the data
In order to do comparative analysis between countries, global population data is needed, which wasn't originally included.  To do this, another file is read in that includes the population data, which is then joined with the global dataset on 
```{r add_pop_global}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region), 
        sep = ", ",
        na.rm=TRUE,
        remove = FALSE)

#Add population, need to grab data from site
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

#Join with global dataset
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```


### Analyzing US data by each state, and then as a whole
``` {r analyze_US_by_state}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/ Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

```

```{r analyze_US_totals}
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/ Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

```


## Visualizing Data

First examine US cases vs. deaths over time
``` {r visuzlize_US_data}
US_totals %>%
  filter(cases >0) %>%
  ggplot(aes(x = date, y = cases))+
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title = "COVID19 in US", y = NULL)

```


Examining individual states.  New York is used here, but any valid state can be input and visualized
``` {r visuzlize_indiv_US_state}
state <- "New York"
US_by_state %>%
  filter(Province_State == state)%>%
  filter(cases >0) %>%
  ggplot(aes(x = date, y = cases))+
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title = str_c("COVID19 in ", state), y = NULL)
```

Looking at new cases vs. new deaths to gain a better understanding
``` {r further_transform_data}
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
    new_deaths = deaths - lag(deaths))
    
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
    new_deaths = deaths - lag(deaths))

```

``` {r graph_US_new_cases}
US_totals %>%
  ggplot(aes(x = date, y = new_cases))+
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US", y = NULL)


```


``` {r graph_new_state}
state <- "Ohio"
US_by_state %>%
  filter(Province_State == state)%>%
  filter(cases >0) %>%
  ggplot(aes(x = date, y = new_cases))+
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title = str_c("COVID19 in ", state), y = NULL)

```


```{r analyze_US_state_totals}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000*cases/population,
            deaths_per_thou = 1000*deaths/population) %>%
  filter(cases > 0, population >0)

```

Looking at deaths per thousand, may be better for comparative analysis between states since their populations can vary greatly
```{r find_deaths_per_thou}
US_state_totals %>%
  slice_min(deaths_per_thou, n = 10)

US_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

```


#Modeling Data

Creating model to make predictions on deaths per thousand and then graphing the results to compare predictions vs actual outcomes
``` {r linear_model}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

US_state_totals %>% slice_max(cases_per_thou)
US_state_totals %>% slice_min(cases_per_thou)

x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)

#Add pred column to US_state_totals
US_state_totals %>% mutate(pred = predict(mod))


#new dataset
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
```


```{r plot_acutal_v_predict}
US_tot_w_pred %>%
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```
